Diverse Genotype-by-Weather Interactions in Switchgrass

### 2024-06-26 R/SI_Appendix_Datasets.Rmd
revision2: Drop MO from Midwest GR50 mash analysis due to GWAS pop structure issues

numSNPs_v <- c(19000, 19000, 19000, 19000, 24000, 33000)
suffix_v <- c("FL50_Gulf_and_Midwest", "FL50_Gulf", "GR50_Gulf_and_Midwest",
              "GR50_Gulf", "GR50_Midwest", "FL50_Midwest")
```

@@

    if (i == 2 & j == 1) {
      betas <- readRDS(file = here("analysis", "mash", popkey[[i]], "inputs", 
                                 paste0("B_hat_random_df_24000topSNPs_",
                                        phekey[[j]], "_", popkey[[i]], ".rds")))
      stderr <- readRDS(file = here("analysis", "mash", popkey[[i]], "inputs", 
                                 paste0("S_hat_random_df_24000topSNPs_",
                                        phekey[[j]], "_", popkey[[i]], ".rds")))
    } else if (i == 2 & j == 2) {
      betas <- readRDS(file = here("analysis", "mash", popkey[[i]], "inputs", 
                                 paste0("B_hat_random_df_33000topSNPs_",
                                        phekey[[j]], "_", popkey[[i]], ".rds")))
      stderr <- readRDS(file = here("analysis", "mash", popkey[[i]], "inputs", 
                                 paste0("S_hat_random_df_33000topSNPs_",
                                        phekey[[j]], "_", popkey[[i]], ".rds")))
    } else {
      betas <- readRDS(file = here("analysis", "mash", popkey[[i]], "inputs", 
                                 paste0("B_hat_random_df_19000topSNPs_",
                                        phekey[[j]], "_", popkey[[i]], ".rds")))
      stderr <- readRDS(file = here("analysis", "mash", popkey[[i]], "inputs", 
                                 paste0("S_hat_random_df_19000topSNPs_",
                                        phekey[[j]], "_", popkey[[i]], ".rds")))
    }
    print(paste(names(popkey)[i], names(phekey)[j], "random SNP set (input into greedy mash algorithm) has", nrow(betas), "SNPs"))
  }
}

@@

```{r}
switchgrassGWAS:::get_marker_df(m_gulf_gr_hyp) # 45931
switchgrassGWAS:::get_marker_df(m_gulf_fl_hyp) # 50490
switchgrassGWAS:::get_marker_df(m_midw_gr_hyp) # 49030
switchgrassGWAS:::get_marker_df(m_midw_fl_hyp) # 63367
switchgrassGWAS:::get_marker_df(m_gam_gr_hyp)  # 44765
switchgrassGWAS:::get_marker_df(m_gam_fl_hyp)  # 35735
```

```{r}
m <- m_gulf_gr_hyp
dataset_creation_function <- function(m, filename){

@@

2N 59.5Mb: biggest effect at TX2, mostly opposite to other gardens or other gardens have NSE
2N 60.9Mb: biggest effect at TX2, AP between TX1 and TX2
2 Gulf effects I would say look nothing like that.
```{r}
overlap <- dataset_8 %>%
  select(mash_idx, mash_subpop, everything()) |> 
  distinct(mash_idx, .keep_all = TRUE)
overlap
```

>>>>

numSNPs_v <- c(19000, 19000, 19000, 19000, 33000, 33000)
suffix_v <- c("FL50_Gulf_and_Midwest", "FL50_Gulf", "GR50_Gulf_and_Midwest",
              "GR50_Gulf", "GR50_Midwest", "FL50_Midwest")
```

@@

    if (i == 2 & j == 1) {
      betas <- readRDS(file = here("analysis", "mash", popkey[[i]], "inputs", 
                                 paste0("B_hat_random_df_33000topSNPs_",
                                        phekey[[j]], "_", popkey[[i]], ".rds")))
      stderr <- readRDS(file = here("analysis", "mash", popkey[[i]], "inputs", 
                                 paste0("S_hat_random_df_33000topSNPs_",
                                        phekey[[j]], "_", popkey[[i]], ".rds")))
    } else if (i == 2 & j == 2) {
      betas <- readRDS(file = here("analysis", "mash", popkey[[i]], "inputs", 
                                 paste0("B_hat_random_df_33000topSNPs_",
                                        phekey[[j]], "_", popkey[[i]], ".rds")))
      stderr <- readRDS(file = here("analysis", "mash", popkey[[i]], "inputs", 
                                 paste0("S_hat_random_df_33000topSNPs_",
                                        phekey[[j]], "_", popkey[[i]], ".rds")))
    } else {
      betas <- readRDS(file = here("analysis", "mash", popkey[[i]], "inputs", 
                                 paste0("B_hat_random_df_19000topSNPs_",
                                        phekey[[j]], "_", popkey[[i]], ".rds")))
      stderr <- readRDS(file = here("analysis", "mash", popkey[[i]], "inputs", 
                                 paste0("S_hat_random_df_19000topSNPs_",
                                        phekey[[j]], "_", popkey[[i]], ".rds")))
    }
    print(paste(names(popkey)[i], names(phekey)[j], "random SNP set (input into greedy mash algorithm) has", nrow(betas), "SNPs"))
  }
}

@@

```{r}
switchgrassGWAS:::get_marker_df(m_gulf_gr_hyp) # 45931
switchgrassGWAS:::get_marker_df(m_gulf_fl_hyp) # 50490
switchgrassGWAS:::get_marker_df(m_midw_gr_hyp) # 57729
switchgrassGWAS:::get_marker_df(m_midw_fl_hyp) # 63367
switchgrassGWAS:::get_marker_df(m_gam_gr_hyp)  # 44765
switchgrassGWAS:::get_marker_df(m_gam_fl_hyp)  # 35735
```

```{r}
m <- m_midw_gr_hyp
dataset_creation_function <- function(m, filename){

@@

2N 59.5Mb: biggest effect at TX2, mostly opposite to other gardens or other gardens have NSE
2N 60.9Mb: biggest effect at TX2, AP between TX1 and TX2
2 Gulf effects I would say look nothing like that.
```{r}
overlap <- dataset_8 %>%
  dplyr::select(mash_idx, mash_subpop, everything()) |> 
  distinct(mash_idx, .keep_all = TRUE)
overlap
```

### 2024-06-26 manuscript/revision2/SI_Appendix/SI_Appendix.qmd
revision2: Drop MO from Midwest GR50 mash analysis due to GWAS pop structure issues

![Types of GxE present between of pairs of jointly re-estimated effects in eight common gardens, for effects with lfsr \< 0.05 at both gardens in each pair. Effect patterns: Sign: Effects that differ in sign at these pairs of gardens (p \< 0.05, lfsr). Magnitude: Effects identical in sign (p \< 0.05, lfsr) that differ in magnitude by a factor of \>0.4. Not Distinguishable: Effects not distinguishable by magnitude nor sign of the effect, with no measurable GxE. b) The number of effects with each GxE type for the onset of vegetative growth (green-up date) within and between two genetic subpopulations.](images/S2_Greenup.png)

@@

![Types of GxE present between of pairs of jointly re-estimated effects in eight common gardens, for effects with lfsr \< 0.05 at both gardens in each pair. Effect patterns: Sign: Effects that differ in sign at these pairs of gardens (p \< 0.05, lfsr). Magnitude: Effects identical in sign (p \< 0.05, lfsr) that differ in magnitude by a factor of \>0.4. Not Distinguishable: Effects not distinguishable by magnitude nor sign of the effect, with no measurable GxE. b) The number of effects with each GxE type for the onset of reproductive growth (green-up date) within and between two genetic subpopulations.](images/S3_Flowering.png)

@@

![All QTL identified in the outbred mapping population displayed GxE, in that all QTL had effects that differed in at least two gardens. Gardens are on the y-axis in latitudinal order; the effects of the significant QTL in Figure 4 are displayed as differences between allelic effects for the two sets of grandparent lines in the outbred mapping panel. ](images/Figure_S4_QTL_effects_by_trait.png){#fig-qtl-by-env}

@@

| Phenology type | Subpopulation | Effects Selected per GWAS, Strong Subset | Effects Selected per GWAS, Random Subset | Total Markers, Strong Subset | Total Markers, Random Subset |
|:-----------|-----------:|-----------:|-----------:|------------|------------|
| Greenup Date   |          Gulf |                                    19000 |                                    38000 | 45931                        | 91862                        |
|                |       Midwest |                                    24000 |                                    48000 | 49030                        | 98060                        |
|                |          Both |                                    19000 |                                    38000 | 44765                        | 89530                        |
| Flowering Date |          Gulf |                                    19000 |                                    38000 | 50490                        | 100980                       |
|                |       Midwest |                                    33000 |                                    66000 | 63367                        | 126734                       |
|                |          Both |                                    19000 |                                    38000 | 35735                        | 71470                        |

@@

| Greedy Algorithm Iteration \# |     Gulf Greenup Date |           Midwest Greenup Date |    Both Greenup Date | Gulf Flowering Date     | Midwest Flowering Date         | Both Flowering Date     |
|:----------|----------:|----------:|----------:|-----------|-----------|-----------|
| 1                             |              Identity |                       Identity |             Identity | Identity                | Identity                       | Identity                |
| 2                             |         Simple Het. 2 |               Single Effect MO |     Single Effect MI | Simple Het. 1           | Simple Het. 1                  | Single Effect TX3       |
| 3                             |      Single Effect MO |               Single Effect NE |     Single Effect SD | Single Effect MI        | Single Effect MI               | Single Effect TX1       |
| 4                             |      Single Effect NE |              Single Effect TX2 |     Single Effect NE | Single Effect TX3       | Single Effect TX2              | Single Effect OK        |
| 5                             | Daylength (14d prior) |              Single Effect TX1 |    Single Effect TX2 | Single Effect SD        | Single Effect MO               | Single Effect MI        |
| 6                             |      Single Effect MI |               Single Effect MI |     Single Effect OK | Single Effect OK        | Single Effect TX3              | Single Effect NE        |
| 7                             |      Single Effect SD |               Single Effect OK |    Single Effect TX3 | Single Effect NE        | Daylength change (s), 2d prior | Single Effect TX2       |
| 8                             |     Single Effect TX2 |              Single Effect TX3 |        Simple Het. 1 | Single Effect TX1       | Single Effect SD               | Single Effect MO        |
| 9                             |    Cumulative GDD, 6d |                  Simple Het. 1 |    Single Effect TX1 | Single Effect TX2       | \-                             | Single Effect SD        |
| 10                            |     Single Effect TX3 | Daylength change (s), 7d prior |     Single Effect MO | Single Effect MO        | \-                             | Simple Het. 1           |
| 11                            |   Data-driven $PCA_2$ |                             \- |         Temp Ave, 1d | Cumulative rainfall, 7d | \-                             | Cumulative rainfall, 1d |
| 12                            |                    \- |                             \- | Daylength change (s) | \-                      | \-                             | \-                      |

>>>>

![Types of GxE present between of pairs of jointly re-estimated effects in eight common gardens, for effects with lfsr \< 0.05 at both gardens in each pair. Effect patterns: Sign: Effects that differ in sign at these pairs of gardens (p \< 0.05, lfsr). Magnitude: Effects identical in sign (p \< 0.05, lfsr) that differ in magnitude by a factor of \>0.4. Not Distinguishable: Effects not distinguishable by magnitude nor sign of the effect, with no measurable GxE. b) The number of effects with each GxE type for the onset of vegetative growth (green-up date) within and between two genetic subpopulations.](images/Fig_S2.png)

@@

![Types of GxE present between of pairs of jointly re-estimated effects in eight common gardens, for effects with lfsr \< 0.05 at both gardens in each pair. Effect patterns: Sign: Effects that differ in sign at these pairs of gardens (p \< 0.05, lfsr). Magnitude: Effects identical in sign (p \< 0.05, lfsr) that differ in magnitude by a factor of \>0.4. Not Distinguishable: Effects not distinguishable by magnitude nor sign of the effect, with no measurable GxE. b) The number of effects with each GxE type for the onset of reproductive growth (green-up date) within and between two genetic subpopulations.](images/Fig_S3.png)

@@

![All QTL identified in the outbred mapping population displayed GxE, in that all QTL had effects that differed in at least two gardens. Gardens are on the y-axis in latitudinal order; the effects of the significant QTL in Figure 4 are displayed as differences between allelic effects for the two sets of grandparent lines in the outbred mapping panel.](images/Figure_S4_QTL_effects_by_trait.png){#fig-qtl-by-env}

@@

| Phenology type | Subpopulation | Effects Selected per GWAS, Strong Subset | Effects Selected per GWAS, Random Subset | Total Markers, Strong Subset | Total Markers, Random Subset |
|:-----------|-----------:|-----------:|-----------:|------------|------------|
| Greenup Date   |          Gulf |                                    19000 |                                    38000 | 45931                        | 91862                        |
|                |       Midwest |                                    33000 |                                    66000 | 57729                        | 115458                       |
|                |          Both |                                    19000 |                                    38000 | 44765                        | 89530                        |
| Flowering Date |          Gulf |                                    19000 |                                    38000 | 50490                        | 100980                       |
|                |       Midwest |                                    33000 |                                    66000 | 63367                        | 126734                       |
|                |          Both |                                    19000 |                                    38000 | 35735                        | 71470                        |

@@

| Greedy Algorithm Iteration \# |     Gulf Greenup Date |     Midwest Greenup Date |    Both Greenup Date | Gulf Flowering Date     | Midwest Flowering Date         | Both Flowering Date     |
|:----------|----------:|----------:|----------:|-----------|-----------|-----------|
| 1                             |              Identity |                 Identity |             Identity | Identity                | Identity                       | Identity                |
| 2                             |         Simple Het. 2 |         Single Effect OK |     Single Effect MI | Simple Het. 1           | Simple Het. 1                  | Single Effect TX3       |
| 3                             |      Single Effect MO |            Equal Effects |     Single Effect SD | Single Effect MI        | Single Effect MI               | Single Effect TX1       |
| 4                             |      Single Effect NE |        Single Effect TX3 |     Single Effect NE | Single Effect TX3       | Single Effect TX2              | Single Effect OK        |
| 5                             | Daylength (14d prior) |        Single Effect TX2 |    Single Effect TX2 | Single Effect SD        | Single Effect MO               | Single Effect MI        |
| 6                             |      Single Effect MI |  Cumulative GDD, 12C, 1d |     Single Effect OK | Single Effect OK        | Single Effect TX3              | Single Effect NE        |
| 7                             |      Single Effect SD | Cumulative GDD, 12C, 28d |    Single Effect TX3 | Single Effect NE        | Daylength change (s), 2d prior | Single Effect TX2       |
| 8                             |     Single Effect TX2 |                       \- |        Simple Het. 1 | Single Effect TX1       | Single Effect SD               | Single Effect MO        |
| 9                             |    Cumulative GDD, 6d |                       \- |    Single Effect TX1 | Single Effect TX2       | \-                             | Single Effect SD        |
| 10                            |     Single Effect TX3 |                       \- |     Single Effect MO | Single Effect MO        | \-                             | Simple Het. 1           |
| 11                            |   Data-driven $PCA_2$ |                       \- |         Temp Ave, 1d | Cumulative rainfall, 7d | \-                             | Cumulative rainfall, 1d |
| 12                            |                    \- |                       \- | Daylength change (s) | \-                      | \-                             | \-                      |

### 2024-06-26 manuscript/revision2/revision2.qmd
revision2: Drop MO from Midwest GR50 mash analysis due to GWAS pop structure issues

Eight GxWeather covariance structures were selected by the greedy algorithm, one to two per subpopulation: five for green-up date, and three for flowering date (SI Appendix, Table S2). Of these eight, five had mass on them in *mash* models of the strong effects (@fig-covar B,D). Three of these five matrices had negative covariances between sites between Texas and North gardens (@fig-covar A), while two had all positive or near-zero covariances.

For green-up date, SNP-associated phenotypic effects covaried with different weather-based cues in the Gulf & in Both subpopulations (@fig-covar B). In total, 65% of the posterior weight of strong SNP effects in the *mash* model of Gulf green-up fell on a covariance matrix constructed using the covariance of day length 14 days prior to the date of green-up. The covariance matrix for this weather cue was visually similar to the observed pattern of phenotypic correlation for the timing of vegetative growth in the Gulf subpopulation (@fig-map B; @fig-covar A). *Mash* models of the timing of vegetative growth in the Midwest subpopulation and Both subpopulations did not include this GxWeather covariance; the Midwest had no weight on any GxWeather covariance, while Both subpopulations had non-zero weights on two additional GxWeather covariance types, average temperature one day prior to green-up, and the day length change in seconds in the day prior to green-up (@fig-covar B). The average temperature covariance matrix had negative covariances between Texas and North gardens, though not as strong as the negative phenotypic correlations seen in Both subpopulations (@fig-map B; @fig-covar A). Only the Gulf subpopulation and Both subpopulations had mass on any GxWeather covariance matrices (@fig-covar C) for vegetative growth.

@@

We expected to observe common rank-changing GxE at the level of individual loci as this is a key theoretical prediction of local adaptation [@felsenstein1976theoretical; @levene1953genetic; @kawecki2004conceptual; @hedrick1986genetic]. To determine the frequency of rank-changing GxE, we used the local false sign rate (lfsr), an analogue of the lfdr that establishes confidence in the effect sign, not the effect's difference from zero, to determine significance. We required lfsr significance (p \< 0.05) in both gardens to include effects. This means that our tests for a sign change between gardens carry an equal statistical burden to those for effects with the same sign. We separated kinds of effects at the level of individual loci into SNP effects that differ in sign between gardens (effects with rank-changing GxE between gardens), SNP effects that differ in magnitude (effects that are large in one garden or region and smaller in others), and SNP effects that are indistinguishable in two gardens (@fig-effects A).

For green-up date for the Gulf subpopulation, hundreds to thousands of pairwise effects exhibited rank-changing GxE, or a difference in sign between pairs of Texas and North gardens (@fig-effects B; SI Appendix, Fig S2). 78.7% of pairwise comparisons between North and Texas gardens had a difference in sign, while only 28.6% and 0.2% of North-North or Texas-Texas comparisons had a difference in sign, respectively (@fig-effects B). The majority of pairwise effects for greenup for the Midwest (\>55%) and Both (\>85%) subpopulations were the same sign, and effects most often differed in magnitude between gardens within and between the regions (@fig-effects B; SI Appendix, Fig S2).

For flowering date for the Gulf subpopulation, less than 2% of pairwise effects exhibited rank-changing GxE, or a difference in effect sign within or between regions (@fig-effects B). More effects differed in magnitude between the Texas and North regions than within these regions (42.7% vs \<20%; @fig-effects B), while the majority of effects had no GxE. The Midwest population had relatively few significant effects for flowering, but a large proportion of these differed in sign between Texas and North regions (42.7%) or within the North region (65.4%). Finally, in Both subpopulations, less than 20% of pairwise effects differed in sign (@fig-effects B). Most differences in sign were between TX1, the southernmost garden, and all other gardens (SI Appendix, Figure S3). Similarly, more effects that differed in magnitude included gardens in the Texas region (52.3-55.9%), and most effect pairs in the North region were not distinguishable (91.5%).

@@

Our analysis of the timing of vegetative growth in the Gulf and in Both subpopulations revealed many alleles with rank-changing GxE, or sign changes in their effects between the Texas and North gardens (@fig-effects B). As phenological timings are major components of plant fitness, this result supports theoretical models that local adaptation should involve antagonistic pleiotropy at the level of individual loci [@kawecki2004conceptual; @felsenstein1976theoretical; @levene1953genetic; @hedrick1986genetic]. In addition, it advances local adaptation research by using more than two field sites and a wide range of genetic variation to determine that rank-changing GxE for phenological timings is common in small genomic regions [@wadgymar2017identifying; @savolainen2013ecological; @lowry2019qtl].

@@

![Example hypothesis-driven covariance matrices specified in *mash* and the posterior weights placed on all covariance matrices. (a) Common gardens are arranged in latitudinal order within the matrices. Top row: Four example canonical covariance matrices. Canonical matrices (purple) have simple interpretations, such as equal effects across all common gardens, or effects specific to a single common garden. Bottom row: Five example GxWeather covariance matrices specified for the green-up date or flowering date phenotype; these matrices were created from environment-specific correlations across eight common gardens, and are described in Table 1. (b,d) Total posterior weight placed on each covariance matrix type specified for (b) green-up date and (d) flowering date *mash* models, within and between two genetic subpopulations. Covariance matrices included in *mash* that had zero posterior weight in all three *mash* runs on the genetic subpopulations, such as the identity matrix, are not shown. (c,e) Total posterior weight placed on covariance matrices that were hypothesized or canonical, for the (c) green-up date phenotype and (e) flowering date phenotype.](images/Figure_2.png){#fig-covar}

![Types of GxE present between of pairs of jointly re-estimated effects in eight common gardens, for effects with lfsr \< 0.05. a) Examples of effect patterns at three pairs of sites with three types of GxE. All effects are for an alternate allele, with the reference allele effect defined to be zero at both gardens and represented by the dashed vertical line. Sign: Effects that differ in sign at these pairs of gardens (p \< 0.05, lfsr). Magnitude: Effects identical in sign (p \< 0.05, lfsr) that differ in magnitude by a factor of \>0.4. Not Distinguishable: Effects not distinguishable by magnitude nor sign of the effect, with no measurable GxE. b) The fraction of effects with each GxE type for the onset of vegetative growth (green-up date) and reproductive growth (flowering date), within and between two genetic subpopulations. Common gardens are grouped by the larger region they came from: North gardens are within the natural range of the Midwest subpopulation, and include MO, NE, MI, and SD, while Texas gardens are within the natural range of the Gulf subpopulation, and include TX1, TX2, and TX3.](images/Figure_3_GxE.png){#fig-effects}

![Overlaps of QTL from an outbred pseudo-F2 cross and with jointly re-estimated SNP effects in the 1% tail of significance from a diversity panel. Dotted lines indicate permutation-based significance thresholds for each weather-related function. Stars indicate QTL with significant enrichment for SNPs in the 1% *mash* tail; G, M, and B indicate which subpopulation had enrichment: G - Gulf subpopulation, M -Midwest subpopulation, B - both subpopulations. Rug plots show genomic locations of SNPs in the 1% *mash* tail for flowering date for each subpopulation: Both subpopulations are above the plot panel, the Gulf subpopulation is above the x-axis, and the Midwest subpopulation is below the x-axis. a) QTL mapping for the onset of vegetative growth (green-up date), and three weather-related functions of green-up date. b) QTL mapping for the onset of reproductive growth (flowering date), and two weather-related functions of flowering date.](images/Figure_4_QTL_Overlaps.png){#fig-qtl}

>>>>

Nine GxWeather covariance structures were selected by the greedy algorithm, one to two per subpopulation: six for green-up date, and three for flowering date (SI Appendix, Table S2). Of these nine, six had mass (\>0.1%) on them in *mash* models of the strong effects (@fig-covar B,D). Five of these six matrices had negative covariances between gardens in the Texas and North regions (@fig-covar A), while one had all positive or near-zero covariances.

For green-up date, SNP-associated phenotypic effects covaried with different weather-based cues in the Gulf & in Both subpopulations (@fig-covar B). In total, 65% of the posterior weight of strong SNP effects in the *mash* model of Gulf green-up fell on a covariance matrix constructed using the covariance of day length 14 days prior to the date of green-up. The covariance matrix for this weather cue was visually similar to the observed pattern of phenotypic correlation for the timing of vegetative growth in the Gulf subpopulation (@fig-map B; @fig-covar A). *Mash* models of the timing of vegetative growth in the Midwest subpopulation and Both subpopulations did not include this GxWeather covariance. The Midwest had 0.15% weight on a GxWeather covariance matrix of cumulative GDD for the 28d prior to green-up, a GxWeather matrix that was visually similar to the Midwest's observed pattern of phenotypic correlation (@fig-map B). Both subpopulations had non-zero weights on two additional GxWeather covariance types, average temperature one day prior to green-up, and the day length change in seconds in the day prior to green-up (@fig-covar B). The average temperature covariance matrix had negative covariances between Texas and North gardens, though not as strong as the negative phenotypic correlations seen in Both subpopulations (@fig-map B; @fig-covar A). The Gulf subpopulation and Both subpopulations had substantially more mass on GxWeather covariance matrices than the Midwest population (@fig-covar C) for the onset of vegetative growth.

@@

We expected to observe common rank-changing GxE at the level of individual loci as this is a key theoretical prediction of local adaptation [@felsenstein1976theoretical; @levene1953genetic; @kawecki2004conceptual; @hedrick1986genetic]. To determine the frequency of rank-changing GxE, we used the local false sign rate (lfsr), an analogue of the lfdr that establishes confidence in the effect sign, not the effect's difference from zero, to determine significance. We required lfsr significance (p \< 0.05) in both gardens to include effects. This means that our tests for a sign change between gardens carry an equal statistical burden to those for effects with the same sign. We separated kinds of effects at the level of individual loci into SNP effects that differ in sign between gardens (effects with rank-changing GxE between gardens), SNP effects that differ in magnitude (effects that are large in one garden or region and smaller in others), and SNP effects that are indistinguishable in two gardens (similarly large or small in comparisons between gardens) (@fig-effects A).

For green-up date for the Gulf subpopulation, hundreds to thousands of pairwise effects exhibited rank-changing GxE, or a difference in sign between pairs of Texas and North gardens (@fig-effects B; SI Appendix, Fig S2). 78.7% of pairwise comparisons between North and Texas gardens had a difference in sign, while only 28.6% and 0.2% of North-North or Texas-Texas comparisons had a difference in sign, respectively (@fig-effects B). The majority of pairwise effects for greenup for the Midwest (\>73%; 68-342 effects) were indistinguishable, or had no GxE. The majority of pairwise effects for Both (\>85%) subpopulations were the same sign, and effects most often differed in magnitude between gardens within and between the regions; however, these differences in effect sign and magnitude were mostly between the MO site and other sites (@fig-effects B; SI Appendix, Fig S2).

For flowering date for the Gulf subpopulation, less than 2% of pairwise effects exhibited rank-changing GxE, or a difference in effect sign within or between regions (@fig-effects B). More effects differed in magnitude between the Texas and North regions than within these regions (42.7% vs \<20%; @fig-effects B), while the majority of effects had no GxE. The Midwest population had relatively few significant effects for flowering (0-174 per pairwise comparison), but a large proportion of these differed in sign between Texas and North regions (42.7%) or within the North region (65.4%). Finally, in Both subpopulations, less than 20% of pairwise effects differed in sign (@fig-effects B). Most differences in sign were between TX1, the southernmost garden, and all other gardens (SI Appendix, Figure S3). Similarly, more effects that differed in magnitude included gardens in the Texas region (52.3-55.9%), and most effect pairs in the North region were not distinguishable (91.5%).

@@

Our analysis of the timing of vegetative and reproductive growth revealed many alleles with rank-changing GxE, or sign changes in their effects between the Texas and North gardens (@fig-effects B). The Gulf & Both subpopulations had rank-changing GxE for green-up date, while the Midwest subpopulation had rank-changing GxE for flowering date. As phenological timings are major components of plant fitness, this result supports theoretical models that local adaptation should involve antagonistic pleiotropy at the level of individual loci [@kawecki2004conceptual; @felsenstein1976theoretical; @levene1953genetic; @hedrick1986genetic]. In addition, it advances local adaptation research by using more than two field sites and a wide range of genetic variation to determine that rank-changing GxE for phenological timings is common in small genomic regions [@wadgymar2017identifying; @savolainen2013ecological; @lowry2019qtl].

@@

![Example hypothesis-driven covariance matrices specified in *mash* and the posterior weights placed on all covariance matrices. (a) Common gardens are arranged in latitudinal order within the matrices. Top row: Four example canonical covariance matrices. Canonical matrices (purple) have simple interpretations, such as equal effects across all common gardens, or effects specific to a single common garden. Bottom row: Five example GxWeather covariance matrices specified for the green-up date or flowering date phenotype; these matrices were created from environment-specific correlations across eight common gardens, and are described in Table 1. (b,d) Total posterior weight placed on each covariance matrix type specified for (b) green-up date and (d) flowering date *mash* models, within and between two genetic subpopulations. Covariance matrices included in *mash* that had zero posterior weight in all three *mash* runs on the genetic subpopulations, such as the identity matrix, are not shown. (c,e) Total posterior weight placed on covariance matrices that were hypothesized or canonical, for the (c) green-up date phenotype and (e) flowering date phenotype.](images/Figure_2_six_sites_Midwest.png){#fig-covar}

![Types of GxE present between of pairs of jointly re-estimated effects in eight common gardens, for effects with lfsr \< 0.05. a) Examples of effect patterns at three pairs of sites with three types of GxE. All effects are for an alternate allele, with the reference allele effect defined to be zero at both gardens and represented by the dashed vertical line. Sign: Effects that differ in sign at these pairs of gardens (p \< 0.05, lfsr). Magnitude: Effects identical in sign (p \< 0.05, lfsr) that differ in magnitude by a factor of \>0.4. Not Distinguishable: Effects not distinguishable by magnitude nor sign of the effect, with no measurable GxE. b) The fraction of effects with each GxE type for the onset of vegetative growth (green-up date) and reproductive growth (flowering date), within and between two genetic subpopulations. Common gardens are grouped by the larger region they came from: North gardens are within the natural range of the Midwest subpopulation, and include MO, NE, MI, and SD, while Texas gardens are within the natural range of the Gulf subpopulation, and include TX1, TX2, and TX3.](images/Figure_3_Fraction_GxE_by_Region_six_midwest_sites.png){#fig-effects}

![Overlaps of QTL from an outbred pseudo-F2 cross and with jointly re-estimated SNP effects in the 1% tail of significance from a diversity panel. Dotted lines indicate permutation-based significance thresholds for each weather-related function. Stars indicate QTL with significant enrichment for SNPs in the 1% *mash* tail; G, M, and B indicate which subpopulation had enrichment: G - Gulf subpopulation, M -Midwest subpopulation, B - both subpopulations. Rug plots show genomic locations of SNPs in the 1% *mash* tail for flowering date for each subpopulation: Both subpopulations are above the plot panel, the Gulf subpopulation is above the x-axis, and the Midwest subpopulation is below the x-axis. a) QTL mapping for the onset of vegetative growth (green-up date), and three weather-related functions of green-up date. b) QTL mapping for the onset of reproductive growth (flowering date), and two weather-related functions of flowering date.](images/Figure_4_QTL_Overlaps_6_Midwest_Gardens.png){#fig-qtl}
